---
title: "US Census code demo"
author: "Peter S. Larson"
date: "2021/11/30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## US Census data

The US Census and its associated data collection efforts are an excellent resource for background data for your public health studies. 

In this code set I will be showing you how to access US Census data through R.

First, you need to get an api key from the US census site.

Go to this site: http://api.census.gov/data/key_signup.html

Sign up for the key. Wait for it to arrive. Don't forget to activate it!


```{r census}
## load necessary packages
library(tidycensus)
library(tidyverse)
library(tmap)
library(dplyr)

## put census api key
### individual key can be obtained from http://api.census.gov/data/key_signup.html

## Here is where you put you newly obtained api key.
#census_api_key("011780b73f165bdca5122d9829bd7749ef2203e6", install=TRUE, overwrite = TRUE)

## Make a basic graphic to show the syntax 
age10 <- get_decennial(geography = "state", 
                       variables = "P013001", 
                       year = 2010)

head(age10)

## Let's make a figure of ages by state
age10 %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point() +
  ggtitle("Age by state") +
  xlab("Age") +
  ylab("State")

```

Now let's find some variables we are interested in.


```{r how_to_find_variables}


## We can find variables of interest by looking in the survey product of interest and then specifying a year (or whatever else is necessary)

## get variable from ACS in year 2019
m19 <- load_variables(2019, "acs5", cache = TRUE)

## find appropriate variable from above (median income)
#B06011_001

## median household income data from the 2015-2019 ACS for tract in Michigan
mi_medincome <-
  get_acs(
    geography = "tract",
    variables = c(medincome = "B06011_001"),
    state = "MI",
    year = 2019
  )  %>%
  dplyr::rename(median_income = estimate) %>%
  dplyr::select(GEOID, median_income)

## find appropriate variable from above (income in the past 12 months below poverty level)
#B10059_002

## income in the past 12 months below poverty level data from the 2015-2019 ACS for tract in Michigan
mi_poverty <-
  get_acs(
    geography = "tract",
    variables = c(poverty = "B10059_002"),
    state = "MI",
    year = 2019
  )  %>%
  dplyr::rename(poverty = estimate) %>%
  dplyr::select(GEOID, poverty)

## find appropriate variable from above (mean age by sex)
#B01002_001

## income in the past 12 months below poverty level data from the 2015-2019 ACS for tract in Michigan
mi_mean_age <-
  get_acs(
    geography = "tract",
    variables = c(mean_age = "B01002_001"),
    state = "MI",
    year = 2019
  )  %>%
  dplyr::rename(mean_age = estimate) %>%
  dplyr::select(GEOID, mean_age)

## get variable from Census in year 2020.We have to use the data choice pl because it is the only one that is avaiable for 2020.
m20 <- load_variables(2020, "pl", cache = TRUE)

## find appropriate variable from above (black or african american race)
#P1_004N

## black/african american race data from the decennial for tract in Michigan
mi_black <-
  get_decennial(
    geography = "tract",
    variables = c(blackrace = "P1_004N"),
    state = "MI",
    year = 2020
  ) %>%
  dplyr::rename(black_pop = value) %>%
  dplyr::select(GEOID, black_pop)

## Put all this in a single file
final_data <- mi_black %>%
  dplyr::left_join(mi_mean_age) %>%
  dplyr::left_join(mi_poverty) %>%
  dplyr::left_join(mi_medincome)

```
  
Making maps with the census api is really, really easy. We just specify "geometry = TRUE" and it automatically gives us a spatial object (sf) to work with. 

``` {r making_maps}
### Thematic Map

## Income
mi_income <- get_acs(
  geography = "tract",
  variables = "B06011_001",
  state = "MI",
  geometry = TRUE
)

## It's an sf object so we can use dplyr code.
mi_income <-  mi_income %>%
  dplyr::rename(median_income = estimate) %>%
  dplyr::mutate(median_income = replace_na(median_income, 0)) %>%
  sf::st_make_valid()

#We can now make a map
tm_shape(mi_income) +
  tm_polygons("median_income",title = "Median Income", lwd = 0.01, border.alpha = 0.1) +
  tm_legend(title = "Median Income", title.position = c("right", "top")) +
  tm_layout (bg.color = "white")


tm_shape(sf::st_make_valid(mi_income)) +
  tm_polygons(
    "median_income",
    title = "Median Income ($)",
    lwd = 0.01,
    border.alpha = 0.1
  ) +
  tm_shape(mi_income) +
  tm_polygons(
    "moe",
    title = "Margin of error",
    lwd = 0.01,
    border.alpha = 0.1,
    alpha = .2
  ) 

```

Let's make a poverty map

```{r poverty_map}

## Poverty
mi_poverty2 <- get_acs(
  geography = "tract",
  variables = "B10059_002",
  state = "MI",
  geometry = TRUE
) %>%
  sf::st_make_valid()

mi_poverty2 <- mi_poverty2 %>%
  dplyr::rename(poverty = estimate) %>%
  dplyr::mutate(poverty = replace_na(poverty, 0)) 
#Map
tm_shape(mi_poverty2) + tm_polygons(
  "poverty",
  title = "Homes in Poverty (%)",
  palette = "YlGnBu",
  lwd = 0.2,
  border.alpha = 0.5
) + 
  tm_legend(title = "Homes in Poverty", title.position = c("right", "top"))

```

Age map

```{r age_map}


## Mean Age
mi_mean_age2 <- get_acs(
  geography = "tract",
  variables = "B01002_001",
  state = "MI",
  geometry = TRUE
) %>%
  sf::st_make_valid()

mi_mean_age2 <- mi_mean_age2 %>%
  dplyr::rename(age = estimate) %>%
  dplyr::mutate(age = replace_na(age, 0))

tm_shape(mi_mean_age2) + tm_polygons(
  "age",
  title = "Mean Age (years)",
  palette = "YlGn",
  lwd = 0.2,
  border.alpha = 0.5
) + tm_legend(title = "Mean Age", title.position = c("right", "top"))
```


```{r black_aa_map}

## black/african american
# mi_black2<- get_decennial(geography = "tract",
#                           variables = "P1_004N",
#                           state = "MI",
#                           geometry = TRUE) %>%
#   sf::st_make_valid()
# 
# mi_black2 <- mi_black2 %>%
#   dplyr::rename(black = estimate)%>% 
#   dplyr::mutate(black = replace_na(black, 0))
# 
# tm_shape(mi_black2) + tm_polygons("black", title = "Black/African American Population (%)", palette = "PuBuGn", lwd = 0.2) + tm_legend(title ="Black/African American Population", title.position = c("right", "top"))

```