---
title: "SUD code demo"
author: "Peter S. Larson"
date: "2021/11/30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals

Using the data set for substance abuse centers, Michigan hospitals, Michigan roads and Michigan counties, perform the following operations:

1. Find the number of substance abuse centers in each Michigan county. Add this variable to the data frame for Michigan counties. Create a map of number of substance abuse centers by county using tmap.

2. Find the distance of each substance abuse center to the nearest hospital and add this variable to the data frame of substance abuse centers. Produce a histogram of the distance of each substance abuse center to the nearest hospital. Create a map of substance abuse centers where the size of the point represents the distance to the nearest hospital. 

3. Create a buffer around each substance abuse center of 15 miles. Count up the number of hospitals within each buffer. Create a map of buffers where the color of the buffers represents the number of hospitals within each buffer. Do the same for a buffer of 5 miles.

Make sure to include titles and legends in your maps. Submit your work in PDF format also including your code in a separate R script.

```{r get_data}
library(readr)
library(readxl)
library(dplyr)
library(janitor)
library(sf)
library(spData)
library(raster)

# example of reading in csv data
# hf_phc <- hf %>% 
#   dplyr::filter(facility_type == "primary health care facility") %>%
#   st_as_sf(coords = c("longitude","latitude"),crs = '+proj=longlat +datum=WGS84 +no_defs',agr="constant")

## Alternatively you can use this from the sp package
## Set coordinates for each 
# library(sp)
# coordinates(hf_tertiary) <- ~longitude + latitude
# coordinates(hf_phc) <- ~longitude +latitude

## Get Michigan counties
michigan_counties <-
  raster::shapefile("Michigan_SUD_code/GIS_data/Counties/Counties_(v17a).shp")

## Convert to sf (if you want to)
michigan_counties <- st_as_sf(michigan_counties)
#michigan_outline <- st_as_sf(michigan_outline)

## Get SUD centers 
SUD_centers <-
  raster::shapefile("Michigan_SUD_code/GIS_data/SUD_data/Substance_Use_Disorder_Care.shp")
## Convert to sf (if you want to)
SUD_centers <- st_as_sf(SUD_centers)

## Get hospitals 
hospitals <- raster::shapefile("Michigan_SUD_code/GIS_data/health_care/Health_Care.shp")
## Convert to sf (if you want to)
hospitals <- st_as_sf(hospitals) %>%
  dplyr::filter(Facility_1 == "Hospital")

# ### Dplyr ####
# mtcars_new <- mtcars %>%
#   dplyr::filter(mpg <20) %>%
#   dplyr::select(mpg, cyl, wt) %>%
#   dplyr::mutate(new_variable = mpg*2)


## Alternatively, you can use subset with a normal SpatialPointsDataFrame

## Make a map to see if this all makes sense.
library(tmap)

tm_shape(michigan_counties) +
  tm_polygons(col = "grey90") +
  tm_shape(SUD_centers) +
  tm_dots(size = .25, col = "red", alpha=.35) +
    tm_shape(hospitals) +
  tm_dots(size = .15, col = "black", alpha=.5) +
  tm_layout(
    main.title = "SUD centers and hospitals in Michigan",
    bg.color = "grey70",
    fontface = "plain",
    fontfamily = "serif"
  ) +
  tm_add_legend(
    'symbol',
    col = c("red", "black"),
    border.col = "grey40",
    border.lwd = 1,
    size = c(.5, .5),
    labels = c('SUD',"Hospitals"),
    title = "Legend"
  )

```

1. Find the number of substance abuse centers in each Michigan county. Add this variable to the data frame for Michigan counties. Create a map of the number of substance abuse centers by county using tmap.

```{r SUD_centers_by_county}

## Count up number of SUD centers by county and add to the Michigan counties layer (sf method)
michigan_counties$SUD_count <-
  lengths(st_intersects(michigan_counties, SUD_centers))

## Create a map of number of substance abuse centers by county using tmap.
tm_shape(michigan_counties) +
  tm_polygons("SUD_count",
              style = "quantile",
              title = "Number of SUD centers",
              alpha = .5) +
  tm_layout(
    main.title = "Number of SUD centers by county",
    bg.color = "grey70",
    fontface = "plain",
    fontfamily = "serif"
  )

```
2. Find the distance of each substance abuse center to the nearest hospital and add this variable to the data frame of substance abuse centers. Produce a histogram of the distance of each substance abuse center to the nearest hospital. Create a map of substance abuse centers where the size of the point represents the distance to the nearest hospital.

```{r distance_to_nearest_hospital}
library(GISTools)
library(ggplot2)

## Get the distances from each of the SUD_centers to each of the hospitals
distances <- st_distance(SUD_centers, hospitals) * 1609.34

## Find the distance of each substance abuse center to the nearest hospital and add this variable to the data frame of substance abuse centers.
SUD_centers$distance_to_hospital <- ft2miles(as.vector(apply(distances,1, min)))

#Produce a histogram of the distance of each substance abuse center to the nearest hospital.
ggplot(data=SUD_centers, aes(x=distance_to_hospital)) +
  geom_histogram(fill="#69b3a2", color="#e9ecef")+
  ggtitle("Distribution of distances to nearest hospital to each SUD center") +
  xlab("Distance to nearest hospital")

#Create a map of substance abuse centers where the size of the point represents the distance to the nearest hospital.
tm_shape(michigan_counties) +
  tm_polygons() +
  tm_shape(SUD_centers) +
  tm_bubbles("distance_to_hospital",
             style = "quantile") +
  tm_layout(
    main.title = "Number of SUD centers by county in Michigan",
    bg.color = "grey90",
    fontface = "plain",
    fontfamily = "serif",
    legend.outside = TRUE
  )

```

3. Create a buffer around each substance abuse center of 15 miles. Count up the number of hospitals within each buffer. Create a map of buffers where the color of the buffers represents the number of hospitals within each buffer. Do the same for a buffer of 5 miles.

```{r buffers}

## Create a buffer around each substance abuse center of 15 miles. 

st_crs(SUD_centers) # Metres
## If necessary, you can change the projection using st_transform(). See documentation
sud_buf <- st_buffer(SUD_centers, 15*1609.34) # it appears the map is in meters, so I converted 15 miles to meters.

## Let's look at the buffers in relation to hospitals really quick
tm_shape(michigan_counties) +
  tm_polygons() +
  tm_shape(sud_buf) +
  tm_borders() +
  tm_shape(hospitals) +
  tm_dots(size=.25) +
  tm_layout(
    main.title = "Hospitals and 15 miles buffers of SUD centers",
    bg.color = "grey90",
    fontface = "plain",
    fontfamily = "serif"
  )

## Count up the number of hospitals within each buffer. 
sud_buf$hospital_count_15_miles <-
  lengths(st_intersects(sud_buf, hospitals))

## Create a map of buffers where the color of the buffers represents the number of hospitals within each buffer.
tm_shape(michigan_counties) +
  tm_polygons() +
  tm_shape(sud_buf) +
  tm_polygons("hospital_count_15_miles", style = "quantile", alpha=.5, legend.hist = TRUE) +
  tm_layout(
    main.title = "Num hospitals within 15 mi of SUD center by county",
    bg.color = "grey70",
    fontface = "plain",
    fontfamily = "serif",
    legend.outside = TRUE
  )

```



