---
title: "Zoom_7_code_exercise"
author: "Peter S. Larson"
date: "2021/12/7"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial data extraction

In the last exercise, we merged data based on common identifiers. The data all came from the same geometries (counties.)

But what if we have data where there aren't common identifiers? 

In this case, we perform our merges based on spatial locations. To show how this is done, we will use a data set of malaria tests, where the location of the home is known. 

```{r data_read_in}
# Load in the malaria case data
ETH_malaria_data <- read.csv("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week1/Lab_files/Data/mal_data_eth_2009_no_dups.csv", header=T) %>% # Case data
  dplyr::mutate(parasite_presence = ifelse(pf_pos > 0, "Positive", "Negative")) ## Create a binary outcome.

## Set coordinates 
coordinates(ETH_malaria_data) <- ~longitude+latitude

## convert to sf object
ETH_malaria_data <- st_as_sf(ETH_malaria_data)

## Get pops
pops <- read_csv("GIS_data/Ethiopia/eht_pops.csv")

## Get some admin boundaries and population at the same time. 
ETH_Adm_2 <- st_as_sf(raster::shapefile("GIS_data/Ethiopia/Ethiopia_adm2_uscb_2020.shp")) %>% # get admin boundaries and make it an sf object at the same time
  left_join(pops %>% dplyr::select(GEO_MATCH, BTOTL_2000)) %>% #join the population data to this using GEO_MATCH as a common identifier
  dplyr::rename(pops = BTOTL_2000) ## rename BTOTL_2000 to pops, just ot make it easier for us to use

## Set these so they are the same crs, really not necessary to make a map, but necessary for the data extraction step.
st_crs(ETH_malaria_data) <- st_crs(ETH_Adm_2)

## Make a map to show we aren't crazy
tm_shape(ETH_Adm_2) +
  tm_polygons("pops", title = "Population", palette= get_brewer_pal(palette="Greys", n=5, plot=FALSE), alpha=.5) +
  tm_shape(ETH_malaria_data) +
  tm_symbols(
    "parasite_presence",
    size = .25,
    border.col = "black",
    border.lwd = 1,
    palette = c("white", "red"),
    title.size = "Parasite presence"
  ) +
  tm_layout(main.title = "Malaria data set")

```

## Covariates

Remember, we want to try and test for associations between spatial risk factors and disease. 

To do this, we have to get some data from other GIS layers.

We've already got data for population from the polygon data. Let's also get a line file for rivers and an elevation raster.

```{r get_other_layers}

## Elevation raster
elev <- raster::raster("GIS_data/Ethiopia/Elevation/ETH_msk_alt.gri")

## Rivers 
rivers <- raster::shapefile("GIS_data/Ethiopia/Rivers/ETH_water_lines_dcw.shp")
rivers <- st_as_sf(rivers)

## Let's plot them so we can convince ourselves that we have the correct data
tm_shape(elev) +
  tm_raster(
    title = "Elevation",
    palette = get_brewer_pal(palette = "Greys", n = 5, plot = FALSE),
    legend.hist = TRUE
  ) +
  tm_shape(rivers, alpha = .5) +
  tm_lines("gray39") +
  tm_layout(main.title = "Elevation and rivers in Ethiopia", legend.outside = TRUE)

```

## Data extraction

Now let's do some data extraction 


```{r data_extraction}

## Population

## Get population numbers for each point, actually get all of the data from the population/adm 2 data set. We can do this using st_intersection()
ETH_malaria_data <- st_intersection(ETH_Adm_2, ETH_malaria_data)

## Let's look at it to see if it worked.
head(ETH_malaria_data)

## Looks like it did. Now let's get the other variables.

## Elevation 
ETH_malaria_data$elevation <- raster::extract(elev, ETH_malaria_data)

## Get distance to river
distances <- st_distance(ETH_malaria_data, rivers) * 0.001
ETH_malaria_data$distance_to_river <- (as.vector(apply(distances,1, min)))

## Now let's look at a summary of the data
summary(ETH_malaria_data)

```


## Models 

Our goal here is to test for associations between presence of malaria and population, elevation and distance to river. 

We can do that using a regression model as before. Plus, we'll make some cool plots.

```{r models}

## Our outcome is binary, which means we can't use a regular scatterplot. We'll get a little fancy. 

## First we have to convert the parasite presence to a number 
ETH_malaria_data <- ETH_malaria_data %>%
  dplyr::mutate(parasite_01 = ifelse(parasite_presence=="Positive", 1, 0))

ggplot(data=ETH_malaria_data, aes(x=pops, y=parasite_01)) +
  geom_point()+
  geom_smooth() +
  ggtitle("Malaria and population") +
  xlab("Population") +
  ylab("Parasite presence")

ggplot(data=ETH_malaria_data, aes(x=elevation, y=parasite_01)) +
  geom_point()+
  geom_smooth() +
  ggtitle("Malaria and elevation") +
  xlab("Elevation") +
  ylab("Parasite presence")

ggplot(data=ETH_malaria_data, aes(x=distance_to_river, y=parasite_01)) +
  geom_point()+
  geom_smooth() +
  ggtitle("Malaria and distance to river") +
  xlab("Distance to river") +
  ylab("Parasite presence")

### We'll also use compareGroups to get a nice table
table_malaria <- compareGroups(parasite_presence ~ pops + elevation + distance_to_river, data=as.data.frame(ETH_malaria_data))
table_malaria <- createTable(table_malaria, show.p.overall = FALSE, show.ratio = TRUE, show.all=TRUE)
export2md(table_malaria)

## Not very convincing, but we'll make a model anyway
fit_malaria <-
  glm(
    parasite_01 ~  pops + elevation + distance_to_river ,
    data = ETH_malaria_data,
    family = "binomial"
  )
summary(fit_malaria)
par(mfrow=c(2,2))
plot(fit_malaria)
par(mfrow=c(1,1))

## Add back to our data set
ETH_malaria_data$predictions <- exp(predict(fit_malaria))/(1+exp(predict(fit_malaria)))
ETH_malaria_data$residuals <- fit_malaria$residuals - min(fit_malaria$residuals)

## Make a couple of maps
predictions <- tm_shape(ETH_Adm_2) +
  tm_polygons(col = "white") +
  tm_shape(ETH_malaria_data) +
  tm_bubbles("predictions", style = "quantile", scale = .75) +
  tm_layout(main.title = "Probability of malaria")

residuals <- tm_shape(ETH_Adm_2) +
  tm_polygons(col = "white") +
  tm_shape(ETH_malaria_data) +
  tm_bubbles("residuals") +
  tm_layout(main.title = "Residuals")

current.mode <- tmap_mode("plot")
tmap_arrange(predictions, residuals, widths = c(.5, .5))
tmap_mode(current.mode)


```